---
editor_options: 
  chunk_output_type: console
---

# Results: Occupancy predictors

## Prepare libraries

```{r load_libs_results02, eval=FALSE}
# to load data
library(readxl)

# to handle data
library(dplyr)
library(readr)
library(forcats)
library(tidyr)
library(purrr)
library(stringr)

# to wrangle models
source("code/fun_model_estimate_collection.r")
source("code/fun_make_resp_data.r")

# plotting
library(ggplot2)
library(patchwork)
source('code/fun_plot_interaction.r')
```

## Get data from hypothesis sheet

```{r eval=FALSE}
# read in the excel sheet containing information on the best supported hypothesis
sheet_names <- readxl::excel_sheets("data/results/all_hypoComparisons_allScales.xlsx")
which_sheet <- which(str_detect(sheet_names, "Best"))

hypothesis_data <- readxl::read_excel("data/results/all_hypoComparisons_allScales.xlsx",
                                      sheet = sheet_names[which_sheet])

# Subsetting the data needed to call in each species' model coefficient information
hypothesis_data <- select(hypothesis_data,
                          Scientific_name, Common_name,
                          contains("Best supported hypothesis"))

# pivot longer
hypothesis_data <- pivot_longer(hypothesis_data,
                                cols = contains("Best"),
                                names_to = "scale", values_to = "hypothesis")
# fix scale to numeric
hypothesis_data <- mutate(hypothesis_data,
                          scale = if_else(str_detect(scale, "10"), "10km", "2.5km"))

# list the supported hypotheses
# first separate the hypotheses into two columns
hypothesis_data <- separate(hypothesis_data, col = hypothesis, sep = "; ", 
                            into = c("hypothesis_01", "hypothesis_02"),
                            fill = "right") %>% 
  # then get the data into long format
  pivot_longer(cols = c("hypothesis_01","hypothesis_02"),
               values_to = "hypothesis") %>% 
  # remove NA where there is only one hypothesis
  drop_na() %>% 
  # remove the name column
  select(-name)

# correct the name landCover to lc
hypothesis_data <- mutate(hypothesis_data,
                          hypothesis = replace(hypothesis,
                                               hypothesis %in% c("landCover", "climate",
                                                                 "elevation"), 
                                               c("lc","clim","elev")))
```

## Read model estimates

```{r read_model_estimates, eval=FALSE}
# which file to read model estimates from
hypothesis_data <- mutate(hypothesis_data,
                          file_read = glue::glue('data/results/results_model_est_{scale}/{hypothesis}_modelEst.xlsx'))

# read in data as list column
model_data <- mutate(hypothesis_data,
                     model_est = map2(file_read, Scientific_name, function(fr, sn){
                       readxl::read_excel(fr, sheet = sn)
                     }))

# rename model data components and separate predictors
names <- c("predictor", "coefficient", "se", "ci_lower", "ci_higher", "z_value", "p_value")

# get data for plotting: separate the interaction terms and make the response df
model_data <- mutate(model_data, 
                     model_est = map(model_est, function(df){
  colnames(df) <- names
  df <- separate_interaction_terms(df)
  df <- make_response_data(df)
  return(df)
}))

# keep significant predictors
model_data <- model_data[map_int(model_data$model_est, nrow) > 0,]

# remove filename
model_data <- select(model_data, -file_read) %>% 
  unnest(model_est) %>% 
  unnest(data)
```

Elevation always varies between 0 to 2625m
Landcover predictors are proportions - so 0 to 1
But climate varies:

prec_interannual: 0 to 300
temp_interannual: 0 to 50
bio_17: 0 to 80
bio_18: 0 to 300

## Prepare interaction plots

Select useful columns and split by species, scale, and hypothesis.

```{r prep_figures_interaction, eval=FALSE}
# select useful columns
plot_data <- select(model_data, 
                   Scientific_name, Common_name,
                   scale, hypothesis,
                   predictor, modulator, m_group, seq_x, mean, ci)

# nest at the plot level and ungroup
plot_data <- group_by(plot_data, 
                      Scientific_name, Common_name,
                      scale, hypothesis,
                      predictor, modulator) %>%
  nest() %>% 
  ungroup()

# make plots
plot_data$plot <- pmap(select(plot_data, data, predictor, modulator), 
                      make_interaction_plot)

# over all b map make interaction plo
plot_data <- select(plot_data, 
                   Common_name, Scientific_name, scale, hypothesis, plot)

plot_data <- group_by(plot_data, Common_name, Scientific_name, scale, hypothesis) %>% 
  summarise(figures = list(plot))
```

```{r}
fig_list <- pmap(plot_data, function(Common_name, Scientific_name, scale, hypothesis, figures){
  final_plot <- wrap_plots(figures, ncol = 7)+
    patchwork::plot_annotation(
      title = glue::glue('{Common_name} ({Scientific_name})'),
      subtitle = glue::glue('scale: {scale}, best supported hypothesis: {hypothesis}'),
      tag_levels = "a",
      tag_prefix = "(", tag_suffix = ")",
      theme = theme(plot.title = element_text(size = 10, face = "bold")))
  
  return(final_plot)
})
```

```{r make_interactions_pdf, eval=FALSE}
# make pdf of patchwork
cairo_pdf(file = "figs/fig_occupancy.pdf", width = 10, height = 3,
          onefile = TRUE)
{
  print(fig_list)
}
dev.off()
```

## Show example interaction

```{r show_one_species, eval=TRUE}
library(magick)
fig_interaction = image_read_pdf("figs/fig_interaction_clim_elev.pdf", pages = 31)
fig_interaction
```

