---
editor_options: 
  chunk_output_type: console
---

# Predicting species-specific occupancy as a function of significant predictors

This script plots species-specific probabilities of occupancy as a function of significant environmental predictors as well as coefficient plots. 

## Prepare libraries

```{r load_libs_results02}
# to load data
library(readxl)
# to handle data
library(dplyr)
library(readr)
library(forcats)
library(tidyr)
library(purrr)
library(stringr)
# plotting
library(ggplot2)
library(patchwork)
```

## Read data

```{r}
# read data
data <- read_csv("data/results/data_predictor_effect.csv")
```

# Plot prob of occupancy as a function of environmental predictors

```{r echo=FALSE}

fig_occupancy <- map2(
  data$scientific_name,
  function(df, scientific_name) {
    ggplot(df) +
      geom_line(aes(coefficient, se,
        col = predictor
      )) +
      coord_cartesian(
        xlim = c(0, 1),
        ylim = c(0, 1)
      ) +
      theme_grey(base_size = 6) +
      theme(legend.position = "top") +
      scale_colour_manual(values = pals::kovesi.rainbow(8)) +
      labs(
        x = "scaled predictor value",
        y = "p(occupancy)",
        title = name
      )
  }
)
```

```{r echo=FALSE}
cairo_pdf(
  filename = "figs/fig_occupancy_predictors.pdf",
  onefile = TRUE, width = 6, height = 4
)
fig_occupancy
dev.off()
```


# Create raster surfaces and predict how occupancy varies across an area



