---
editor_options: 
  chunk_output_type: console
---

# Predicting species-specific occupancy as a function of significant predictors

This script plots species-specific probabilities of occupancy as a function of significant environmental predictors as well as coefficient plots. 

## Prepare libraries

```{r load_libs_results02}
# to handle data
library(dplyr)
library(readr)
library(tidyr)
library(purrr)
library(stringr)
library(glue)
# library(data.table)

# nice tables
library(knitr)
library(kableExtra)

# plotting
library(ggplot2)
library(patchwork)
```

## Read data

```{r}
# read coefficient effect data
data <- read_csv("data/results/data_predictor_effect.csv")
```

```{r}
# check for a predictor column
assertthat::assert_that(
  all(c("predictor", "coefficient", "se") %in% colnames(data)),
  msg = "make_response_data: data must have columns called 'predictor',
    'coefficient', and 'se'"
)

# data <- filter(
#   data
#   # p_value <= 0.05 | predictor == "Int"
# )
```

## Prepare predictor data

```{r}
# preparep predictors
predictors = c("bio4a", "bio15a", glue("lc_0{seq(9)}"))

# prepare predictor search strings and scaling power
preds = glue("({predictors})")
preds = str_flatten(preds, collapse = "|")
power = as.numeric(str_extract(data$predictor, "2"))
power[is.na(power)] = 1

# assign predictor name and power
data = mutate(
  data,
  predictor = str_extract(predictor, preds),
  power = power
)
```

## Get predictor responses

```{r}
# make predictor sequences
data = mutate(
  data,
  pred_val = map(
    predictor,
    function(x) { seq(0, 1, 0.05) }
  ),
  # handle squared terms
  pred_val_pow = purrr::map2(
    pred_val, power,
    function(x, y) {x^y}
  )
)

# get coefficient and error times terms
data_resp = mutate(
  data,
  response = map2(
    pred_val_pow, coefficient,
    function(x, y) {x * y}
  ),
  resp_var = map2(
    pred_val_pow, se,
    function(x, y) {x * y}
  )
)
```

## Get probability of occupancy

```{r}
# unnest and get responses
data_resp = unnest(
  data_resp, 
  cols = c("response", "resp_var", "pred_val")
)

# get responses for quadratic terms
data_resp =
  group_by(
  data_resp,
  scientific_name, predictor, pred_val
) %>% 
  select(-power, -coefficient, -se) %>% 
  summarise(
    across(
      .cols = c("response", "resp_var"),
      .fns = sum
    ), 
    .groups = "keep"
  )

# get probability of occupancy
data_resp = ungroup(
  data_resp
) %>% 
  mutate(
    p_occu = 1 / (1 + exp(-response)),
    p_occu_low = 1 / (1 + exp(-(response - resp_var))),
    p_occu_high = 1 / (1 + exp(-(response + resp_var)))
  )
```

## Add scaling for predictors

```{r}
# scale predictors
scale_15a = 40
scale_4a = 0.375

# scale bio4a and bio15a by actual values
data_resp = mutate(
  data_resp,
  pred_val = case_when(
    predictor == "bio4a" ~ pred_val * scale_4a,
    predictor == "bio15a" ~ pred_val * scale_15a,
    T ~ pred_val
  )
)

# make long
data_poccu = select(
  data_resp,
  -response, -resp_var
)
```

```{r}
# select species
soi = c("Culicicapa ceylonensis", "Irena puella", 
        "Copsychus fulicatus", "Pycnonotus jocosus")
which_predictors = c("bio4a")
```

### Figure: Occupancy ~ predictors

```{r}
data_fig = data_poccu %>% 
  filter(
    scientific_name %in% soi,
    predictor %in% which_predictors
  ) |> 
  mutate(
    cat = case_when(
      scientific_name %in% c("Irena puella", "Culicicapa ceylonensis") ~ "forest",
      T ~ "general"
    )
  )

# split data
data_fig = nest(
  data_fig,
  -cat
)
```


```{r}
# make plots
make_occu_fig = function(df, this_fill) {
  ggplot(
    df
  )+
  geom_ribbon(
    aes(
      pred_val,
      ymin = p_occu_low,
      ymax = p_occu_high
    ),
    fill = this_fill,
    alpha = 0.5
  )+
  geom_line(
    aes(
      pred_val, p_occu
    ),
    size = 1
  )+
  facet_grid(
    ~scientific_name
  )+
  theme_test(
    base_family = "Arial"
  )+
  theme(
    strip.text = element_text(
      face = "italic"
    )
  )+
  labs(
   x = "Temperature seasonality",
   y = "Probability of occupancy"
  )
}


fig_occu = map2(data_fig$data, "grey", make_occu_fig)

fig_occu =
  wrap_plots(
  fig_occu[c(2, 1)], ncol = 2
) & plot_annotation(
  tag_levels = "A"
) &
  theme(
    plot.tag = element_text(
      face = "bold"
    )
  )

# save figure
ggsave(
  fig_occu,
  filename = "figs/fig_05.png",
  width = 5, height = 4
)
```

### Figures: Occupancy ~ predictors for all species

```{r}
data_fig = nest(
  data_poccu,
  -scientific_name, -predictor
)

pred_names = c(
  "bio4a" = "Temp. seasonality",
  "bio15a" = "Ppt. seasonality",
  "lc_01" = "Evergreen",
  "lc_02" = "Deciduous",
  "lc_03" = "Mixed/degraded",
  "lc_04" = "Agro./Settled",
  "lc_05" = "Grassland",
  "lc_07" = "Plantation",
  "lc_09" = "Water"
)
pred_names = tibble(
  name = pred_names,
  predictor = names(pred_names)
)
data_fig = left_join(
  data_fig,
  pred_names
)

data_fig = mutate(
  data_fig,
  plots = map(
    data, function(df) {
      ggplot(df)+
      geom_ribbon(
        aes(
          pred_val,
          ymin = p_occu_low,
          ymax = p_occu_high
        ),
        fill = "grey",
        alpha = 0.5
      )+
      geom_line(
        aes(
          pred_val, p_occu
        )
      )+
      coord_cartesian(
        ylim = c(0, 1)
      )+
      theme_test(
        base_family = "Arial"
      )+
      labs(
        x = "Predictor",
        y = "p(Occupancy)"
      )
    }
  )
)

# add names
data_fig = mutate(
  data_fig,
  plots = map2(
    plots, name,
    function(p, name) {
      p = p + labs(
        x = name
      )
    }
  )
)

# summarise as patchwork
data_fig = group_by(
  data_fig,
  scientific_name
) |> 
  summarise(
    plots = list(
      wrap_plots(
        plots,
        ncol = 5
      )
    )
  )

# add title as sp
data_fig = mutate(
  data_fig,
  plots = map2(
    plots, scientific_name,
    function(p, name) {
      p = p & plot_annotation(
        title = name
      )
    }
  )
)
```

```{r}
# save images
cairo_pdf(
  filename = "figs/fig_occupancy_predictors.pdf",
  onefile = TRUE, width = 10, height = 2
)
data_fig$plots
dev.off()
```

## Mapping species occupancy

### Read in raster layers

```{r}
library(terra)
library(sf)
```

```{r}
# read rasters
bio_4a <- rast("data/chelsa/CHELSA_bio10_4a.tif")
bio_15 <- rast("data/chelsa/CHELSA_bio10_15.tif")
landcover = rast("data/landUseClassification/landcover_roy_2015_reclassified.tif")
lc_1km = rast("data/landUseClassification/lc_01000m.tif")
```

### Split landcover into proportions per 1km

```{r}
# separate the fine-scale landcover raster into presence-absence of each class
lc_split = segregate(landcover)

# resample to 1km
# bilinear resampling uses the mean function. 
# mean of N 0s and 1s is the proportion of 1s, ie, proportion of each landcover
lc_split = terra::resample(
  lc_split,
  lc_1km,
  method = "bilinear" 
)

# rename rasters
names(lc_split) = pred_names$name[-c(1,2)]

# save raster of landcover proportion
terra::writeRaster(
  lc_split,
  filename = "data/spatial/raster_landcover_proportion_1km.tif"
)

rm(landcover); gc()
```

```{r eval=FALSE}
# plot proportion of landcover classes
png(width = 1200 * 2, height = 1200 * 2, filename = "figs/fig_landcover_proportion_1km.png", res = 300)
plot(
  lc_split, 
  col = colorspace::sequential_hcl(20, palette = "Viridis"),
  range = c(0, 1)
)
dev.off()
```

### Prepare climatic layers

```{r}
# get ranges
range_4a = terra::minmax(bio_4a)[, 1]
range_15 = terra::minmax(bio_15)[, 1]

# rescale
bio_4a = (bio_4a - min(range_4a)) / (diff(range_4a))
bio_15 = (bio_15 - min(range_15)) / (diff(range_15))

# project to UTM
climate = c(bio_4a, bio_15)
climate = terra::project(
  x = climate, y = lc_1km
)

# make squared terms
climate2 = climate * climate

# add to landcover proportions and plot
landscape = c(climate, lc_split)
```

```{r eval=FALSE}
# plot proportion of landcover classes
png(width = 1200 * 2, height = 1200 * 2, filename = "figs/fig_landscape_1km.png", 
    res = 300)
plot(
  landscape, 
  col = colorspace::sequential_hcl(20, palette = "agSunset", rev = T),
  range = c(0, 1)
)
dev.off()
```

```{r}
# add squared terms
landscape = c(
  climate, climate2, lc_split
)
```

### Prepare soi predictors

Prepare the soi predictor coefficients as a vector of the same length as the number of raster layers.
These will be multiplied with each layer to give the effect of each layer.

```{r}
# get soi coefs
soi_coefs = filter(
  data,
  scientific_name %in% soi
) %>% 
  select(
    -pred_val, -pred_val_pow
  )

# add missing landcover classes
soi_preds = crossing(
  scientific_name = soi,
  predictor = pred_names$predictor,
  power = c(1,2)
)

# remove squared terms for landcover
soi_preds = filter(
  soi_preds,
  !(str_detect(predictor, "lc") & power == 2)
)

soi_coefs = full_join(
  soi_coefs,
  soi_preds
)

# make wide
soi_coefs =
  pivot_wider(
  soi_coefs, 
  id_cols = c("scientific_name"), 
  names_from = c("predictor", "power"), 
  values_from = "coefficient"
)

# get into order
soi_coefs = select(
  soi_coefs,
  scientific_name,
  c(
    "bio4a_1", "bio15a_1",
    "bio4a_2", "bio15a_2"
  ),
  matches("lc")
)

# get vectors of coefficients
soi_coefs = nest(
  soi_coefs,
  -scientific_name
)
```

### Prepare species occupancy

Here, we shall simply multiply each landscape layer with the corresponding predictor coefficient.
Where these are not available, we shall simply multiply the corresponding layer with NA.
The resulting layers will be summed together to get a single response layer, which will then be inverse logit transformed to get the probability of occupancy.

```{r}
# multiply coefficients with layers
soi_occu = map(
  soi_coefs$data,
  .f = function(df) {
    response = unlist(slice(df, 1), use.names = F) * landscape
    response = sum(response, na.rm = TRUE) # remove NA layers, i.e., non-sig preds
    
    # now transform for probability occupancy
    response = 1 / (1 + exp(-response))
  }
)

# assign names
names(soi_occu) = soi_coefs$scientific_name

# make single stack
soi_occu = reduce(soi_occu, c)
names(soi_occu) = str_replace(soi_coefs$scientific_name, " ", "_")
```

```{r}
# use stars for plotting with ggplot
library(stars)
library(colorspace)

soi_occu = st_as_stars(soi_occu)

fig_occu_map = ggplot()+
  geom_stars(
    data = soi_occu
  )+
  scale_fill_binned_sequential(
    palette = "Hawaii",
    name = "p(Occu.)",
    rev = T,
    limits = c(0, 1),
    breaks = seq(0, 1, 0.1), show.limits = T
  )+
  facet_wrap(
    ~band,
    labeller = labeller(
      band = function(x) str_replace(x, "_", " ")
    )
  )+
  coord_sf(
    crs = 32643,
    expand = FALSE
  )+
  theme_test()+
  theme(
    # legend.position = "rg",
    axis.title = element_blank(),
    axis.text = element_blank(),
    legend.key.height = unit(10, "mm"),
    legend.key.width = unit(1, "mm"),
    strip.text = element_text(
      face = "italic"
    ),
    legend.title = element_text(
      vjust = 1
    )
  )
```

```{r}
# save figure
ggsave(
  fig_occu_map,
  filename = "figs/fig_06.png",
  width = 6, height = 6
)
```

